Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})

### 2014-05-17 manuscript.tex
edits to results, tables, figures cont.

The highest frequency of the \emph{Hopscotch} insertion in teosinte was found in \emph{parviglumis} sympatric with cultivated maize. Our initial hypothesis was that the high frequency of the \emph{Hopscotch} element in these populations could be attributed to introgression from maize into teosinte. To investigate this possibility we examined overall patterns of linkage disequilibrium across chromosome one and specifically in the \emph{tb1} region. If the \emph{Hopscotch} is found in these populations due to recent introgression we would expect to find large blocks of linked markers near this element. We find no evidence of elevated linkage disequilibrium between the \emph{Hopscotch} and SNPs surrounding the \emph{tb1} region in our resequenced populations (Fig. \ref{Fig2LD}), and $r^{2}$ in the \emph{tb1} region  does not differ significantly between populations with (average $r^{2}$ of 0.085) and without the \emph{Hopscotch} genotype (average $r^{2}=0.082$). In fact, average $r^{2}$ is lower in the \emph{tb1} region ($r^{2}=0.056$) than across the rest of chromosome 1 ($r^{2}=0.083$) (\ref{Table3R2}). \mbh{LV, please go through and make sure the data entered into all the tables is correct.  In Table3, both sequenced regions were labeled as "Region 1".  I changed the second to Region 2 but don't know if the data in this column are really from Region 2}

% Table 3 $r^{2}$ values
\begin{table}[htbp]
\rowcolors{2}{gray!25}{white}
  \centering
  \caption{$r^{2}$ values between SNPs in the \emph{tb1} region (positions 264,596,664-265,891,456 on chromosome 1 of the maize AGPv2 genome) and the rest of chromosome 1, within the 5' UTR (Sequenced region 1), and within the 66,169 bp upstream region (Sequenced region 2).}
    \begin{tabular}{ccccc}\\\toprule
    \textbf{Population} & \textbf{Chr. 1} & \textbf{\emph{tb1} region} & \textbf{Region 1} & \textbf{Region 2} \\\midrule
    Ejutla A & 0.095 & 0.050 & 0.747 & 0.215 \\
    Ejutla B & 0.069 & 0.051 & 0.660 & 0.186 \\
    La Mesa & 0.070 & 0.053 & 0.914 & 0.766 \\
    San Lorenzo & 0.101 & 0.067 & 0.912 & 0.636 \\
    \bottomrule
    \end{tabular}
  \label{Table3R2}
\end{table}

% Figure 2. LD plots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=150mm]{Fig2LDPlots.pdf}
    \caption{Linkage disequilibrium for SNPs in Mb 261-268 on chromosome 1. The yellow rectangle indicates the location of the \emph{Hopscotch} insertion and the green represents the \emph{tb1} ORF. A) Ejutla A; B) Ejutla B; C) La Mesa; D). San Lorenzo} 
\label{Fig2LD}
  \end{center}
\end{figure*}
%-------------------------------------------------------------------

@@

To assess the contribution of \emph{tb1} to phenotypic variation in tillering in a natural population, we grew plants from seed sampled from the San Lorenzo population of \emph{parviglumis}, which had a high mean frequency (0.44) of the \emph{Hopscotch} insertion from our initial genotyping. We measured tillering index (TI), the ratio of the sum of tiller lengths to plant height, for 216 plants from within the San Lorenzo population, and genotyped plants for the \emph{Hopscotch} insertion. We found the \emph{Hopscotch} segregating at a frequency of 0.65 with no significant deviations from expected frequencies under Hardy-Weinberg equilibrium. After performing a repeated measures ANOVA between our transformed tillering index data and \emph{Hopscotch} genotype we find a weak positive correlation between presence of the \emph{Hopscotch} and tillering index on day 40 (p=0.0848), but no correlation between tillering index and genotype on any other day (\ref{Fig4Boxplots}). Additionally we find no significant correlation between tiller number and \emph{Hopscotch} genotype, or culm diameter and \emph{Hopscotch} genotype in Phenotyping 1. \jri{shouldn't we expect a negative correlation betweeh Hop and TI on day 40? need to have an A and B in the figure and explain one is for pheno1 and one is for pheno2. please explain whiskers and dots on figure too.} \lev{sure, I mean, presumably we would expect things with Hop to have a smaller TI} \jri{yup, but we should mention that the expectation is negative/}

% Figure 4 Phenotyping BoxPlots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=150mm]{Fig4Boxplots.pdf}
    \caption{Box-plot showing tillering index in our greenhouse population from day 20-60. White indicates individuals homozygous for the \emph{Hopscotch}, light grey represents heterozygotes, and dark grey represents homozygotes for the teosinte (No \emph{Hopscotch}) allele. Within boxes, dark black lines represent the median, and the edges of the boxes are the first and third quartiles.} 
\label{Fig4Boxplots}
  \end{center}
\end{figure*}
%-------------------------------------------------------------------
\jri{lots of white space in fig 4 and fig. s1 too.} 
We performed a second grow-out  of teosinte to assess whether lighting conditions or sample size may have affected our ability to detect and effect of \emph{tb1}.  For the second grow-out we measured tillering index every five days through day 50 for 302 plants. We found the \emph{Hopscotch} allele segregating at a frequency of 0.69, \jri{is it in HWE in this pop?} with a 0.6 frequency of \emph{Hopscotch} homozygotes, and a 0.2 frequency of both heterozygotes and homozygotes for the teosinte allele. We found similar patterns, with a weak positive correlation between tillering index and \emph{Hopscotch} genotype at day 40 (p<0.0611), with no significant correlation on any day. Similarly, relationships between \emph{Hopscotch} genotype and tiller number, and \emph{Hopscotch} genotype and culm diameter are not significant.

>>>>

The highest frequency of the \emph{Hopscotch} insertion in teosinte was found in \emph{parviglumis} sympatric with cultivated maize. Our initial hypothesis was that the high frequency of the \emph{Hopscotch} element in these populations could be attributed to introgression from maize into teosinte. To investigate this possibility we examined overall patterns of linkage disequilibrium across chromosome one and specifically in the \emph{tb1} region. If the \emph{Hopscotch} is found in these populations due to recent introgression we would expect to find large blocks of linked markers near this element. We find no evidence of elevated linkage disequilibrium between the \emph{Hopscotch} and SNPs surrounding the \emph{tb1} region in our resequenced populations (Fig. \ref{Fig2LD}), and $r^{2}$ in the \emph{tb1} region  does not differ significantly between populations with (average $r^{2}$ of 0.085) and without (average $r^{2}=0.082$) the \emph{Hopscotch} genotype. In fact, average $r^{2}$ is lower in the \emph{tb1} region ($r^{2}=0.056$) than across the rest of chromosome 1 ($r^{2}=0.083$) (\ref{Table3R2}). \mbh{LV, please go through and make sure the data entered into all the tables is correct.  In Table3, both sequenced regions were labeled as "Region 1".  I changed the second to Region 2 but don't know if the data in this column are really from Region 2}

% Table 3 $r^{2}$ values
\begin{table}[htbp]
\rowcolors{2}{gray!25}{white}
  \centering
  \caption{$r^{2}$ values between SNPs on chromosome 1, in the broad \emph{tb1} region, within the 5' UTR of \emph{tb1} (Region 1), and 66kb upstream of \emph{tb1} (Region 2).}
    \begin{tabular}{ccccc}\\\toprule
    \textbf{Population} & \textbf{Chr. 1} & \textbf{\emph{tb1} region} & \textbf{Region 1} & \textbf{Region 2} \\\midrule
    Ejutla A & 0.095 & 0.050 & 0.747 & 0.215 \\
    Ejutla B & 0.069 & 0.051 & 0.660 & 0.186 \\
    La Mesa & 0.070 & 0.053 & 0.914 & 0.766 \\
    San Lorenzo & 0.101 & 0.067 & 0.912 & 0.636 \\
    \bottomrule
    \end{tabular}
  \label{Table3R2}
\end{table}

% Figure 2. LD plots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=140mm]{Fig2LDPlots.pdf}
    \caption{Linkage disequilibrium for SNPs in Mb 261-268 on chromosome 1. The yellow rectangle indicates the location of the \emph{Hopscotch} insertion and the green represents the \emph{tb1} ORF. A) Ejutla A; B) Ejutla B; C) La Mesa; D). San Lorenzo} 
\label{Fig2LD}
  \end{center}
\end{figure*}
%-------------------------------------------------------------------

@@

To assess the contribution of \emph{tb1} to phenotypic variation in tillering in a natural population, we grew plants from seed sampled from the San Lorenzo population of \emph{parviglumis}, which had a high mean frequency (0.44) of the \emph{Hopscotch} insertion based on our initial genotyping. We measured tillering index (TI), the ratio of the sum of tiller lengths to plant height, for 216 plants (Phenotyping 1) from within the San Lorenzo population, and genotyped plants for the \emph{Hopscotch} insertion. We found the \emph{Hopscotch} segregating at a frequency of 0.65 with no significant deviations from expected frequencies under Hardy-Weinberg equilibrium. After performing a repeated measures ANOVA between our transformed tillering index data and \emph{Hopscotch} genotype we find a weak positive correlation between presence of the \emph{Hopscotch} and tillering index on day 40 (p=0.0848), a result indicating the \emph{Hopscotch} may actually increase tillering in \emph{parviglumis} in contrast to its phenotypic effect in maize. We find no correlation between tillering index and genotype on any other day (\ref{Fig4Boxplots}). Additionally we find no significant correlation between tiller number and \emph{Hopscotch} genotype, or culm diameter and \emph{Hopscotch} genotype in Phenotyping 1.

% Figure 4 Phenotyping BoxPlots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=95mm]{Fig4Boxplots.pdf}
    \caption{Box-plots showing tillering index in greenhouse grow-outs for Phenotyping 1 (A) and Phenotyping 2 (B). White indicates individuals homozygous for the \emph{Hopscotch}, light grey represents heterozygotes, and dark grey represents homozygotes for the teosinte (No \emph{Hopscotch}) allele. Within boxes, dark black lines represent the median, and the edges of the boxes are the first and third quartiles. \jri{please explain whiskers and dots on figure too.}} 
\label{Fig4Boxplots}
  \end{center}
\end{figure*}
%-------------------------------------------------------------------

We performed a second grow-out  of teosinte to assess whether lighting conditions or sample size may have affected our ability to detect and effect of \emph{tb1}.  For the second grow-out we measured tillering index every five days through day 50 for 302 plants. We found the \emph{Hopscotch} allele segregating at a frequency of 0.69, \jri{is it in HWE in this pop?} with a 0.6 frequency of \emph{Hopscotch} homozygotes, and a 0.2 frequency of both heterozygotes and homozygotes for the teosinte allele. We found similar patterns, with a weak positive correlation between tillering index and \emph{Hopscotch} genotype at day 40 (p<0.0611), with no significant correlation on any day. Similarly, relationships between \emph{Hopscotch} genotype and tiller number, and \emph{Hopscotch} genotype and culm diameter are not significant.
